 

### 2021-06-18 vertical-thermal-review.Rmd
Update vertical-thermal-review.Rmd

**Canopy foliage acts as the primary physical barrier between the atmosphere and the forest floor, buffering multiple aspects of the understory conditions. It is critical in influencing -- and is influenced by -- the vertical biophysical gradient (Fig. `r fig_NEON_vertical`).**
 The distribution of this foliage is structured in correspondence with the distribution of tree sizes, and successional stages within a forest, where leaf area density ($LAD$) ie. leaf area per unit volume, increases with tree-height eventually forming the canopy layer.

LAD along the vertical gradient is heterogeneous, with common patterns that vary across forests. Dense canopy forests composed of broadleaf trees can have high $LAD$ in the upper canopy layers, as well as in the understory such as in tropical forests [@ashtonComparisonsStructureMixed1992; @koikeCanopyStructureTropical1993] whereas in temperate forests, although LAD is generally more concentrated in the upper-canopy, undualting patterns are formed by conifers and top vs. bottom heavy foliage trees and sometimes by dense understories [@parkerVerticalProfileCanopy1989; @terborghVerticalComponentPlant1985; @ishiiExploringRelationshipsCanopy2004a; @parkerCanopyLightTransmittance2002]. In open canopy forests, such as needle-leaf forests, LAD can be predominantly accumulated in the lower canopy layers [@lawLeafAreaDistribution2001; @baldocchiSeasonalVariationEnergy1997; @bonan_ecological_2016; @ishiiExploringRelationshipsCanopy2004a]. Soil moisture conditions, topography, gap formations followed by disturbance, all alter foliage patterns and further influence the formation of relatively closed vs. open canopy forests [e.g., @almeidaContrastingFireDamage2016 ; @starkReframingTropicalSavannization2020; @bonan_ecological_2016]. In addition, seasonally dry and wet conditions, deciduousness and phenology contribute to the spatial loss and gain of leaves (Parker & Tibbs 2004; @smithSeasonalDroughtrelatedChanges2019 ; @parkerConsequencesEnvironmentalHeterogeneity2019). In this review, we focus on growing season conditions unless otherwise noted.

@@

A few studies have found $Q_{10}$ to be greater in upper-canopy leaves than in shaded lower-canopy leaves, with significant differences among species [@turnbullScalingFoliarRespiration2003; @harleyEffectsLightTemperature1996], whereas @griffinCanopyPositionAffects2002 found that in Populas deltoides, lower canopy leaf respiration was more sensitive to changes in $T_{air}$ than upper canopy. Other studies have observed no difference in $Q_{10}$ across species vertical profiles and in seedling sun vs. shade leaves [@weerasingheCanopyPositionAffects2014; @bolstad_foliar_1999; @zaragoza-castellsDoesGrowthIrradiance2007; @zaragoza-castellsClimatedependentVariationsLeaf2008]. Although, in @bolstad_foliar_1999, temperate trees showed a slight tendency to increase down the canopy relative to upper-canopy leaves as $T_{air}$ increased, [@bolstad_foliar_1999](*see their Fig. 1*). (*Martijn has some unpublished data on this that we might include.unpublished data showing higher Q10 in shade than in sun leaves for 10 species or so. May make sense to include here: https://github.com/EcoClimLab/vertical-thermal-review/issues/8.*) with evidence of acclimation and a declining trend of respiration with topographical elevation.
Similary, response of respiration to $T_{leaf}$ [partially represented  by $E_{0}$] did not differ within canopy positions of Quercus rubra, whereas on considering site differences, canopy-position effect in respiration was much greater in mesic sites, and seasonal effect was greater in drier sites [@xuSeasonalVariationTemperature2006].

@@

**Vertical gradients affect tree recruitment, growth and survival, and how these respond to climatic variation.**
More interesting in this context is how these shape differences in the climatic sensitivity of growth and mortality. There are numerous demographic differences between understory and canopy trees, which have been linked to both the physical environment and plant traits (e.g., REFS), *and we will not review those here.* *(although maybe we should put one paragraph?)* Demographic differences between canopy trees and understory trees can also influence their their ability to tolerate multiple stressors. Tolerance of more than one biological and biophyical stressors can increase with tree height and light, partly due to non-structural carbon reserve [NSC], with smaller shaded trees being more sensitive due to lower NSC [@niinemetsResponsesForestTrees2010a]. However, growth also declines as trees age. This is partly because autotrophic respiration increases with age that declines NPP, whereas due to increase in morphological and anatomical complexity, GPP also increases [@IncreasingRespiratoryCosts]. Vertical biophysical conditions are bound to have different influenceS on size class and growth rates due to these existing demographic differences.

Greater $VPD$, atmospheric demand and low RH can be more stressful for trees, than soil moisture across biomes-including mesic forests,  as the former limits surface conductance and evapotranspiration to a greater extent than soil moisture [@novickIncreasingImportanceAtmospheric2016; @corlettImpactsWarmingTropical2011; @ruehrWaterAvailabilityDominant2016]. Along with greater VPD, increase in temperature even under adequete water can cause greater tree hydraulic stress due to altered soil-water dynamics than milder drought, in temperate species [@ruehrWaterAvailabilityDominant2016; @parkwilliamsTemperaturePotentDriver2013]. Whereas greater $RH$ can do the opposite, stifling evapotranspiration and thereby increasing $T_{leaf}$ [Fig.3; @perezIncreasingHumidityThreatens2018; @tibbittsHumidityPlants1979]. Altered precipitation regimes cause droughts, that are more  detrimental to larger trees than small trees, because limited soil-water cannot meet the evaporative demand via the long-distance water transport to the canopy [@bennettLargerTreesSuffer2015; @mcgregor_tree_nodate; @bartholomewSmallTropicalForest2020]. However, given adequate moisture to sustain high $g_s$, canopy leaves exposed to more wind and lower humidity should be more effective at cooling when $T_{air}$ exceeds physiological optima (Fig. `r fig_leaf_T`).

@@

Canopies are crucial to microclimatic functioning, as they buffer several aspects of forest microenvironments [Zelleweger et al. 2019]. This buffering is influenced by the hydraulogical-cycle at large [Davis et al. 2019], therefore severe droughts place canopies at risk, particularly as larger trees suffer more during droughts[Bennett et al. 2015]. Further, increasing global temperatures show that canopies of temperate and tropical forests are already functioning close to thermal threshold and beyond photosynthetic optimas [@mau_temperate_2018; @huangAirTemperatureOptima2019]. Such increase in canopy $T_{leaf}$ reduces forest GPP [@pauTropicalForestTemperature2018] in the tropics, where thermal limits are narrower than in temperate forests [Perez and Feeley, 2018].

Canopy structure affects understory species composition and soil fertility, which has been shown to shift under macroclimatic warming and canopy disturbance [@majasalmiImpactTreeCanopy2020; @bertrandCommentForestMicroclimate2020; @maesPlantFunctionalTrait2020]. Macroclimatic warming will disproportionately affect the less warm-adapted functional types within the forest canopy. Even with buffering of climatic extremes, thereby creating holes in the canopy that will take a long time to refill (especially in slow-growing tropical sub-canopy specialists), with associated understory plant thermophilization (death is faster than recruitment and growth) [@duqueThermophilizationAdultJuvenile2015a; @zellweger_forest_2020; @zellweger_seasonal_2019; @defrenneMicroclimateModeratesPlant2013]. This imposes an increasing threat to flora and fauna in microrefugias that are  otherwise buffered from extremes [@scheffersMicrohabitatsReduceAnimal2013].

>>>>

**Canopy foliage acts as the primary physical barrier between the atmosphere and the forest floor, buffering multiple aspects of the understory conditions. It is critical in influencing -- and is influenced by -- the vertical biophysical gradient (Fig. `r fig_NEON_vertical`).**
 The distribution of this foliage is structured in correspondence with the distribution of tree sizes, and successional stages within a forest, where leaf area density (LAD) ie. leaf area per unit volume, increases with tree-height eventually forming the canopy layer.

LAD along the vertical gradient is heterogeneous, with common patterns that vary across forests. Dense canopy forests composed of broadleaf trees can have high LAD in the upper canopy layers, as well as in the understory such as in tropical forests [@ashtonComparisonsStructureMixed1992; @koikeCanopyStructureTropical1993] whereas in temperate forests, although LAD is generally more concentrated in the upper-canopy, undualting patterns are formed by conifers and top vs. bottom heavy foliage trees and sometimes by dense understories [@parkerVerticalProfileCanopy1989; @terborghVerticalComponentPlant1985; @ishiiExploringRelationshipsCanopy2004a; @parkerCanopyLightTransmittance2002]. In open canopy forests, such as needle-leaf forests, LAD can be predominantly accumulated in the lower canopy layers [@lawLeafAreaDistribution2001; @baldocchiSeasonalVariationEnergy1997; @bonan_ecological_2016; @ishiiExploringRelationshipsCanopy2004a]. Soil moisture conditions, topography, gap formations followed by disturbance, all alter foliage patterns and further influence the formation of relatively closed vs. open canopy forests [e.g., @almeidaContrastingFireDamage2016 ; @starkReframingTropicalSavannization2020; @bonan_ecological_2016]. In addition, seasonally dry and wet conditions, deciduousness and phenology contribute to the spatial loss and gain of leaves (Parker & Tibbs 2004; @smithSeasonalDroughtrelatedChanges2019 ; @parkerConsequencesEnvironmentalHeterogeneity2019). In this review, we focus on growing season conditions unless otherwise noted.

@@

A few studies have found $Q_{10}$ to be greater in upper-canopy leaves than in shaded lower-canopy leaves, with significant differences among species [@turnbullScalingFoliarRespiration2003; @harleyEffectsLightTemperature1996], whereas @griffinCanopyPositionAffects2002 found that in Populas deltoides, lower canopy leaf respiration was more sensitive to changes in $T_{air}$ than upper canopy. Other studies have observed no difference in $Q_{10}$ across species vertical profiles and in seedling sun vs. shade leaves [@weerasingheCanopyPositionAffects2014; @bolstad_foliar_1999; @zaragoza-castellsDoesGrowthIrradiance2007; @zaragoza-castellsClimatedependentVariationsLeaf2008]. Although, in @bolstad_foliar_1999, temperate trees showed a slight tendency to increase down the canopy relative to upper-canopy leaves as $T_{air}$ increased, [@bolstad_foliar_1999](*see their Fig. 1*). (*Martijn has some unpublished data on this that we might include.unpublished data showing higher Q10 in shade than in sun leaves for 10 species or so. May make sense to include here: https://github.com/EcoClimLab/vertical-thermal-review/issues/8.*) with evidence of acclimation and a declining trend of respiration with topographical elevation.
Similary, response of respiration to $T_{leaf}$ [partially represented  by $E_{0}$] did not differ within canopy positions of Quercus rubra, whereas in considering site differences, canopy-position effect in respiration was much greater in mesic sites, and seasonal effect was greater in drier sites [@xuSeasonalVariationTemperature2006].

@@

**Vertical gradients affect tree recruitment, growth and survival, and how these respond to climatic variation.**
More interesting in this context is how these shape differences in the climatic sensitivity of growth and mortality. There are numerous demographic differences between understory and canopy trees, which have been linked to both the physical environment and plant traits (e.g., REFS), *and we will not review those here.* *(although maybe we should put one paragraph?)* Demographic differences between canopy trees and understory trees can also influence their their ability to tolerate multiple stressors. Tolerance of more than one biological and biophyical stressors can increase with tree height and light, enabled by non-structural carbon reserve [NSC], with smaller shaded trees being more sensitive due to lower NSC [@niinemetsResponsesForestTrees2010a]. However, growth also declines as trees age. This is partly because autotrophic respiration increases with age that declines NPP, whereas due to increase in morphological and anatomical complexity, GPP also increases [@IncreasingRespiratoryCosts]. Vertical biophysical conditions are bound to influence growth and whole tree functioning.

Greater $VPD$, atmospheric demand and low RH can be more stressful for trees, than soil moisture across biomes-including mesic forests,  as the former limits surface conductance and evapotranspiration to a greater extent than soil moisture [@novickIncreasingImportanceAtmospheric2016; @corlettImpactsWarmingTropical2011; @ruehrWaterAvailabilityDominant2016]. Along with greater VPD, increase in temperature even under adequete water can cause greater tree hydraulic stress due to altered soil-water dynamics than milder drought, in temperate species [@ruehrWaterAvailabilityDominant2016; @parkwilliamsTemperaturePotentDriver2013]. Whereas greater $RH$ can do the opposite, stifling evapotranspiration and thereby increasing $T_{leaf}$ [Fig.3; @perezIncreasingHumidityThreatens2018; @tibbittsHumidityPlants1979]. Altered precipitation regimes can cause droughts, that are more  detrimental to larger trees than small trees, because limited soil-water cannot meet the evaporative demand via the long-distance water transport to the canopy [@bennettLargerTreesSuffer2015; @mcgregor_tree_nodate; @bartholomewSmallTropicalForest2020]. However, given adequate moisture to sustain high $g_s$, canopy leaves exposed to more wind and lower humidity should be more effective at cooling when $T_{air}$ exceeds physiological optima (Fig. `r fig_leaf_T`).

@@

Canopies are crucial to microclimatic functioning, as they buffer several aspects of forest microenvironments [Zelleweger et al. 2019]. This buffering is influenced by the hydraulogical-cycle at large [Davis et al. 2019], therefore severe droughts place canopies at risk, particularly as larger trees suffer more during droughts[Bennett et al. 2015]. Further, increasing global temperatures show that canopies of temperate and tropical forests are already functioning close to thermal threshold and beyond photosynthetic optimas [@mau_temperate_2018; @huangAirTemperatureOptima2019]. This can have implications in the tropics, where thermal limits are narrower than in temperate forests [Perez and Feeley, 2018]. Such increase in canopy $T_{leaf}$ has also shown to reduce forest GPP in the tropics [@pauTropicalForestTemperature2018].

Canopy structure affects understory species composition and soil fertility, which has been shown to shift under macroclimatic warming and canopy disturbance [@majasalmiImpactTreeCanopy2020; @bertrandCommentForestMicroclimate2020; @maesPlantFunctionalTrait2020]. Macroclimatic warming will disproportionately affect the less warm-adapted functional types within the forest canopy, even with buffering of climatic extremes, thereby creating holes in the canopy that will take a long time to refill (especially in slow-growing tropical sub-canopy specialists), with associated understory plant thermophilization (death is faster than recruitment and growth) [@duqueThermophilizationAdultJuvenile2015a; @zellweger_forest_2020; @zellweger_seasonal_2019; @defrenneMicroclimateModeratesPlant2013]. This imposes an increasing threat to the biodiversity in microrefugias that are  otherwise buffered from extremes [@scheffersMicrohabitatsReduceAnimal2013].
